Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-07-07 main.tex
Make minor adjustments.
- Clean up description of strategies to be consistent.
- Explain what training corresponds to for each strategy.
- Add a number of TODOs.

\begin{abstract}
    We present tournament results and several powerful strategies for the Iterated
    Prisoner's Dilemma created using reinforcement learning techniques
    (evolutionary and particle swarm algorithms). These strategies are
    trained to perform well against a corpus of over 100 distinct
    opponents, including many well-known strategies from the literature, and all
    the trained strategies win standard tournaments against the total collection
    of other opponents. We also trained variants to win noisy tournaments.
\end{abstract}

@@

The Axelrod library now contains many parametrised strategies trained using
machine learning
methods. Most are deterministic, use many rounds of memory, and perform
extremely well in tournaments.

@@

The first strategy trained with reinforcement learning in the library is based
on lookup tables. The strategy encodes a set of deterministic responses
based on the opponent's first $n_1$ moves, the opponent's last $m_1$ moves, and
the players last $m_2$ moves. If $n_1 > 0$ then the player has infinite memory
depth, otherwise it has depth $\max{m_1, m_2}$.
This is shown diagrammatically in Figure~\ref{fig:lookerup}.

@@

Although we tried various
combinations of $n_1, m_1,$ and $m_2$, the best performance at the time of
training was obtained for $n_1 = m_1 = m_2 = 2$ and generally for $n_1 > 0$.
The library includes a strategy
called EvolvedLookerUp2\_2\_2 which is among the top strategies in the library.

@@

Gambler is a stochastic variant of LookerUp. Instead of deterministically
encoded moves the lookup table emits probabilities which are
used to choose cooperation or defection.
This is shown diagrammatically in Figure~\ref{fig:gambler}.

@@

The library includes a player trained
with $n_1 = m_1 = m_2 = 2$ that is \emph{mostly deterministic}, with most of the
probabilities being 0 or 1. At one time this strategy outperformed
EvolvedLookerUp2\_2\_2.

This strategy type can be used to train arbitrary memory-$n$ strategies. A
memory one strategy called PSO Gambler Mem 1 was trained, with
probabilities $(\text{Pr}(\text{C}\;|\;\text{CC}),
                \text{Pr}(\text{C}\;|\;\text{CD}),
                \text{Pr}(\text{C}\;|\;\text{DC}),
                \text{Pr}(\text{C}\;|\;\text{DD})) = (1, 0.5217, 0, 0.121)$.
Though it performs well in standard tournaments (see
Table~\ref{tbl:standard_score})
it is not as good as the longer memory strategies, and is bested by a similar
strategy that also uses the first round of play: PSO Gambler 1 1 1.

@@

Strategies based on artificial neural networks can also be trained with an
evolutionary algorithm. A variety of features are computed from the history
of play such as the opponents trailing moves, the total number of cooperations
of the player and the opponent, and several others, which are then input
into a feed forward neural network with one layer and user-supplied width.
This is shown diagrammatically in Figure~\ref{fig:ann}.

@@

@@

We used strategies based on finite state machines to create a number of
strategies. These strategies are deterministic and are efficient computationally.
In each round of play the strategy selects an action based on the current state
and the opponent's last action, transitioning to a new state for the next round.
These strategies can encode a variety
of other strategies, including classic strategies like TitForTat,
handshake strategies, and grudging strategies that always defect after
an opponent defection.
This is shown diagrammatically in Figure~\ref{fig:fsm}.

@@

% TODO: One or more FSM diagrams

@@

We also trained stochastic versions of finite state machine players called
hidden Markov model players or HMMs. Like the finite state machines, these
strategies also encode an internal
state with probabilistic transitions based on the prior round of play to other
states and cooperate or defect with various probabilities at each state. These
are the best performing stochastic strategies in the library but take longer
to train due to their stochasticity.
This is shown diagrammatically in Figure~\ref{fig:hmm}.

@@

@@

Because these strategies inherit many of the properties of the strategies
on which they are based, including using the match length to defect on the last
rounds of play, these strategies were omitted from the tournament results.

\section{Results}

@@

Using the method of fingerprinting described in \cite{ashlock2008fingerprinting}
% TODO This is actually a numerical approximation of the fingerprinting
% described in ashlock's paper. There they consider a fingerprint to be an
% analytic function.
\cite{ashlock2009fingerprint}, we can compare
strategies. For the top performing noisy strategies there is a striking similarity
in the fingerprints.

@@

@@

\section{Methods}

We trained a variety of strategies using evolutionary algorithms, and in the
case of PSO Gambler using a particle swarm algorithm. The evolutionary algorithms
used standard techniques, varying strategies by mutation and crossover, and
evaluating the performance against each opponent for many repetitions. The
best performing strategies in each generation are persisted, variants created,
and objective functions computed again. This process continues for approximately
200 generations or until strategies no longer improve significantly.

@@

All training code is available on github. There are objective functions for
\begin{itemize}
 \item total or mean payoff
 \item total or mean payoff difference (unused in this work)
 \item total Moran process wins (fixation probability)
\end{itemize}

>>>>

\begin{abstract}
    We present tournament results and several powerful strategies for the Iterated
    Prisoner's Dilemma created using reinforcement learning techniques
    (evolutionary and particle swarm algorithms). These strategies are
    trained to perform well against a corpus of over 170 distinct
    opponents, including many well-known strategies from the literature. All
    the trained strategies win standard tournaments against the total collection
    of other opponents. We also trained variants to win noisy tournaments.
\end{abstract}

@@

The Axelrod library now contains many parametrised strategies trained using
machine learning
methods. Most are deterministic, use many rounds of memory, and perform
extremely well in tournaments as will be discussed in Section~\ref{sec:results}.
Training of these strategies will be discussed in Section~\ref{sec:methods}.

These strategies can encode a variety
of other strategies, including classic strategies like Tit For Tat,
handshake strategies, and grudging strategies that always defect after
an opponent defection.

@@

The LookerUp strategy is based on a lookup tables and  encodes a set of
deterministic responses based on the opponent's first $n_1$ moves, the
opponent's last $m_1$ moves, and the players last $m_2$ moves. If $n_1 > 0$ then
the player has infinite memory depth, otherwise it has depth $\max{m_1, m_2}$.
This is illustrated diagrammatically in Figure~\ref{fig:lookerup}.

@@

Training of this strategy corresponds to finding maps from histories to either a
cooperation or a defection.

Although various
combinations of $n_1, m_1,$ and $m_2$ have been tried, the best performance at
the time of
training was obtained for $n_1 = m_1 = m_2 = 2$ and generally for $n_1 > 0$.
A strategy
called EvolvedLookerUp2\_2\_2 is among the top strategies in the library.

@@

Gambler is a stochastic variant of LookerUp. Instead of deterministically
encoded moves the lookup table emits probabilities which are
used to choose cooperation or defection.
This is illustrated diagrammatically in Figure~\ref{fig:gambler}.

@@

Training of this strategy corresponds to finding maps from histories to
a probability of cooperation.

The library includes a strategy trained
with $n_1 = m_1 = m_2 = 2$ that is \emph{mostly deterministic}, with most of the
probabilities being 0 or 1. At one time this strategy outperformed
EvolvedLookerUp2\_2\_2.

This strategy type can be used to train arbitrary memory-$n$ strategies. A
memory one strategy called PSO Gambler Mem 1 was trained, with
probabilities $(\text{Pr}(\text{C}\;|\;\text{CC}),
                \text{Pr}(\text{C}\;|\;\text{CD}),
                \text{Pr}(\text{C}\;|\;\text{DC}),
                \text{Pr}(\text{C}\;|\;\text{DD})) = (1, 0.5217, 0, 0.121)$.
Though it performs well in standard tournaments (see
Table~\ref{tbl:standard_score})
it is not as good as the longer memory strategies, and is bested by a similar
strategy that also uses the first round of play: PSOGambler\_1\_1\_1.

@@

Strategies based on artificial neural networks use a variety of features
computed from the history of play:

\begin{multicols}{2}
    \begin{itemize}
        \item Opponent's first move is C
        \item Opponent's first move is D
        \item Opponent's second move is C
        \item Opponent's second move is D
        \item Player's previous move is C
        \item Player's previous move is D
        \item Player's second previous move is C
        \item Player's second previous move is D
        \item Opponent's previous move is C
        \item Opponent's previous move is D
        \item Opponent's second previous move is C
        \item Opponent's second previous move is D
        \item Total opponent cooperations
        \item Total opponent defections
        \item Total player cooperations
        \item Total player defections
        \item Round number
    \end{itemize}
\end{multicols}

These are then input into a feed forward neural network with one layer and
user-supplied width.  This is illustrated diagrammatically in
Figure~\ref{fig:ann}.

@@

Training of this strategy corresponds to finding parameters of the neural
network.

@@

Strategies based on finite state machines are deterministic and computationally efficient.
In each round of play the strategy selects an action based on the current state
and the opponent's last action, transitioning to a new state for the next round.
This is illustrated diagrammatically in Figure~\ref{fig:fsm}.

@@

Training this strategy corresponds to finding mappings of states and histories
to an action and a state. Figures~\ref{} show two of the trained finite state
machines...
% TODO Update sentence above.
% TODO Add diagrams of FSMs.

@@

A variant of finite state machine strategies are called hidden Markov models
(HMMs). Like the strategies based on finite state machines, these strategies
also encode an internal state however use probabilistic transitions based on the
prior round of play to other states and cooperate or defect with various
probabilities at each state. This is
shown diagrammatically in Figure~\ref{fig:hmm}.

@@

Training this strategy corresponds to finding mappings of states and histories
to probabilities of cooperating as well as probabilities of the next internal
state.

These are the best performing stochastic strategies
in the library but take longer to train due to their stochasticity.

@@

Because these strategies inherit many of the properties of the strategies
on which they are based, including using the match length to defect on the last
rounds of play, not all of these strategies were included in results of this
paper.

\section{Results}\label{sec:results}

@@

Using the method of fingerprinting described in \cite{ashlock2008fingerprinting,
ashlock2009fingerprint}
% TODO This is actually a numerical approximation of the fingerprinting
% described in ashlock's paper. There they consider a fingerprint to be an
% analytic function.
we can compare
strategies. For the top performing noisy strategies there is a striking similarity
in the fingerprints.

@@

% TODO Include collaboration rating heatmaps

@@

% TODO Include collaboration rating heatmaps

\section{Methods}\label{sec:methods}

We trained a variety of strategies using evolutionary algorithms, and in the
case of PSO Gambler using a particle swarm algorithm. The evolutionary algorithms
used standard techniques, varying strategies by mutation and crossover, and
evaluating the performance against each opponent for many repetitions. The
best performing strategies in each generation are persisted, variants created,
and objective functions computed again. This process continues for approximately
200 generations or until strategies no longer improve significantly.
% TODO Improve the explanation here with more details. As described this is not
% reproducible.
% TODO Include references

@@

All training code is available on github. There are objective functions for
% TODO We should archive the code.

\begin{itemize}
 \item total or mean payoff
 \item total or mean payoff difference (unused in this work)
 \item total Moran process wins (fixation probability)
\end{itemize}
